Variance reduction for MCMC methods via martingale representations

### 2020-02-20 aap/Revision/vr-mcmc-revision-new.tex
1st theorem proof update

\begin{proof}%[Proof of Theorem~\ref{prop:29032018a1}]
The expansion obviously holds for $p=1$ and $j=0$.
Indeed, since $\left(\phi_{k}\right)_{k \geq 0}$ is a complete orthonormal system in \(L^2(\mathbb{R}^d, P_{\xi})\), it holds in \(L^2(\mathsf{P})\) that
\[
f(X^x_{1})=\PE[f(X^x_{1})]+\sum_{k\geq1}a_{1,1,k}(x)\phi_{k}(\xi_{1})
\]
with $a_{1,1,k}(x)=\PE[f(X_{1}^x)\phi_{k}(\xi_{1})]$.
Assume now that (\ref{eq:mart_repr})
holds for $p=q$, all $j<q$, and Borel-measurable functions
$f$ with $\mathsf{E}\left[|f(X^x_{q})|^2\right]<\infty$.
Let us prove that the induction assumption holds for $p=q+1$.
Given $f$ with $\mathsf{E}\left[|f(X^x_{q+1})|^2\right]<\infty$,
due to the orthonormality and completeness
of the system $\left(\phi_{k}\right)$, we get that for any $y \in \rset^d$ it holds in \(L^2(\mathsf{P})\) 
\[
f(X^y_{q,q+1}) = \mathsf{E} f(X^y_{q,q+1}) + \sum\limits_{k=1}^{\infty}\mathsf{E}[f(X^y_{q,q+1})\phi_{k}(\xi_{q+1})]\phi_k(\xi_{q+1})
\]
which means that
\begin{equation}
\label{eq:parseval}
\lim_{n \rightarrow \infty}\mathsf{E}[|f(X^y_{q,q+1}) - \mathsf{E} f(X^y_{q,q+1}) - \sum\limits_{k=1}^{n}\mathsf{E}[f(X^y_{q,q+1})\phi_{k}(\xi_{q+1})]\phi_k(\xi_{q+1})|^2] = 0
\end{equation}
Let us define sequence $\psi_{q+1,n}(y) = \mathsf{E}[|f(X^y_{q,q+1}) - \mathsf{E} f(X^y_{q,q+1})|^2] - \sum\limits_{k=1}^{n}\mathsf{E}^2[f(X^y_{q,q+1})\phi_k(\xi_{q+1})]$. Note that $\psi_{q+1,n}(y)$ coincides with the left side of \ref{eq:parseval}, and due to Parseval identity $\Psi_{q+1,n}(y) \rightarrow 0, n \rightarrow \infty$ for any $y \in \rset^d$. Note that $\psi_n(y) < \mathsf{E}[|f(X^y_{q,q+1}) - \mathsf{E} f(X^y_{q,q+1})|^2] \leq \frac{\|f\|^2_{\infty}}{4}$. Hence, by Lebesgue dominated convergence theorem, $\mathsf{E}\psi_n(X_q^x)^2 \rightarrow 0$ and since $\mathsf{E}[f(X^y_{q,q+1})\phi_{k}(\xi_{q+1})]$ is a version of conditional expectation $\mathsf{E}[f(X^x_{q+1}|X^x_q = y]$, it holds in \(L^2(\mathsf{P})\)
\begin{equation}
\label{eq:1-step-decomp}
f(X^x_{q+1}) = \mathsf{E}[f(X^x_{q+1})|\mathcal{G}_k] + \sum\limits_{k=1}^{\infty}a_{q+1,q+1,k}(X^x_{q})\phi_{k}(\xi_{q+1})
\end{equation}
where
\[
a_{q+1,q+1,k}(y) = \mathsf{E}\left[f(X^y_{q,q+1})\phi_k\left(\xi_{q+1}\right)\right]
\]

%we get by
%conditioning on $\mathcal{G}_{q}$,
%\[
%f(X^x_{q+1})=\mathsf{E}\left[\left.f(X^x_{q+1})\right|{\mathcal{G}}_q\right]+\sum_{k\geq1}\alpha_{q+1,q+1,k}\phi_{k}(\xi_{q+1}),
%\]
%where
%\begin{equation*}
%\alpha_{q+1,q+1,k}
%=\mathsf{E}\left[\left.f(X^x_{q+1})\phi_{k}(\xi_{q+1})\right|\mathcal G_q\right].
%\end{equation*}
%By the Markov property of $(X^x_{l})$, it holds $\mathsf{E}[f(X^x_{q+1})|\mathcal{G}_q] = \mathsf{E}[f(X^x_{q+1})| X_q^x]$. Using Markov property together with conditioning on $\mathcal{G}_{q+1}$, we obtain
%\[
%\alpha_{q+1,q+1,k}=\mathsf{E}\left[\left.f(X^x_{q+1})\phi_{k}(\xi_{q+1})\right|X^x_{q}\right]
%=a_{q+1,q+1,k}(X^x_{q})
%\]
%where for $y \in \rset^d$
%\[
%a_{q+1,q+1,k}(y) = \mathsf{E}\left[f(X^y_{q,q+1})\phi_k\left(\xi_{q+1}\right)\right]
%\]

We thus arrive at
\begin{align}
\label{berm:sig_X}
f(X^x_{q+1})=\mathsf{E}\left[\left.f(X^x_{q+1})\right|X^x_{q}\right]+\sum_{k\geq1}a_{q+1,q+1,k}(X^x_{q})\phi_{k}(\xi_{q+1}),
\end{align}
which is the required statement in the case $j=q$.
Now assume $j<q$.
The random variable
$\mathsf{E}\left[\left.f(X^x_{q+1})\right|X^x_{q}\right]$
is square integrable and has the form
$g(X^x_{q})$,
hence the induction hypothesis applies, and we get
\begin{equation}\label{eq:28082017a1}
\mathsf{E}\left[\left.f(X^x_{q+1})\right|X^x_{q}\right]=\mathsf{E}\left[\left.f(X^x_{q+1})\right|X^x_{j}\right]+\sum_{k\geq1}\sum_{l=j+1}^{q}a_{q+1,l,k}(X^x_{l-1})\phi_{k}(\xi_{l})
\end{equation}
with
\begin{align*}
a_{q+1,l,k}(X^x_{l-1}) &= \mathsf{E}\left[\left.\mathsf{E}\left[\left.f(X^x_{q+1})\right|\mathcal{G}_{q}\right]\phi_{k}(\xi_{l})\right|\mathcal{G}_{l-1}\right]
 = \mathsf{E}\left[\left.f(X^x_{q+1})\phi_{k}(\xi_{l})\right|X^x_{l-1}\right].
\end{align*}
where for $y \in \rset^d$,
\[
a_{q+1,l,k}(y) = \mathsf{E}\left[f(X^y_{l-1,q+1})\phi_{k}(\xi_{l})\right]
\]
Formulas \eqref{berm:sig_X}
and~\eqref{eq:28082017a1} conclude the induction step for $p = q+1$ and hence the proof.
\end{proof}

>>>>

\begin{proof}%[Proof of Theorem~\ref{prop:29032018a1}]
The expansion obviously holds for $p=1$ and $j=0$.
Indeed, since $\left(\phi_{k}\right)_{k \geq 0}$ is a complete orthonormal system in \(L^2(\mathbb{R}^d, P_{\xi})\), it holds in \(L^2(\mathsf{P})\) that
f(X^x_{1})=\PE[f(X^x_{1})]+\sum_{k\geq1}a_{1,1,k}(x)\phi_{k}(\xi_{1})
with $a_{1,1,k}(x)=\PE[f(X_{1}^x)\phi_{k}(\xi_{1})]$.
Assume now that (\ref{eq:mart_repr})
holds for $p=q$, all $j<q$, and Borel-measurable functions
$f$ with $\mathsf{E}\left[|f(X^x_{q})|^2\right]<\infty$.
Let us prove that the induction assumption holds for $p=q+1$.
Given $f$ with $\mathsf{E}\left[|f(X^x_{q+1})|^2\right]<\infty$,
due to the orthonormality and completeness
of the system $\left(\phi_{k}\right)$, we get that for any $y \in \rset^d$ it holds in \(L^2(\mathsf{P})\) 
f(X^y_{q,q+1}) = \mathsf{E} f(X^y_{q,q+1}) + \sum\limits_{k=1}^{\infty}\mathsf{E}[f(X^y_{q,q+1})\phi_{k}(\xi_{q+1})]\phi_k(\xi_{q+1})
which means that
\label{eq:parseval}
\lim_{n \rightarrow \infty}\mathsf{E}[|f(X^y_{q,q+1}) - \mathsf{E} f(X^y_{q,q+1}) - \sum\limits_{k=1}^{n}\mathsf{E}[f(X^y_{q,q+1})\phi_{k}(\xi_{q+1})]\phi_k(\xi_{q+1})|^2] = 0
Let us define sequence of functions
\psi_{q+1,n}(y) = \mathsf{E}[|f(X^y_{q,q+1}) - \mathsf{E} f(X^y_{q,q+1})|^2] - \sum\limits_{k=1}^{n}\mathsf{E}^2[f(X^y_{q,q+1})\phi_k(\xi_{q+1})].
which coincide with the left side of \ref{eq:parseval}. Due to Parseval identity $\Psi_{q+1,n}(y) \rightarrow 0, n \rightarrow \infty$ for any $y \in \rset^d$ and $\psi_n(y) < \mathsf{E}[|f(X^y_{q,q+1}) - \mathsf{E} f(X^y_{q,q+1})|^2] \leq \frac{\|f\|^2_{\infty}}{4}$. Hence, by Lebesgue dominated convergence theorem, $\mathsf{E}\psi^2_n(X_q^x) \rightarrow 0$ and since $\mathsf{E}[f(X^y_{q,q+1})]$ is a version of the conditional expectation $\mathsf{E}[f(X^x_{q+1}|X^x_q = y]$, it holds in \(L^2(\mathsf{P})\) that
\label{eq:1-step-decomp}
f(X^x_{q+1}) = \mathsf{E}[f(X^x_{q+1})|X^x_{q+1}] + \sum\limits_{k=1}^{\infty}a_{q+1,q+1,k}(X^x_{q})\phi_{k}(\xi_{q+1})
a_{q+1,q+1,k}(y) = \mathsf{E}\left[f(X^y_{q,q+1})\phi_k\left(\xi_{q+1}\right)\right]
%we get by
%conditioning on $\mathcal{G}_{q}$,
%\[
%f(X^x_{q+1})=\mathsf{E}\left[\left.f(X^x_{q+1})\right|{\mathcal{G}}_q\right]+\sum_{k\geq1}\alpha_{q+1,q+1,k}\phi_{k}(\xi_{q+1}),
%\]
%where
%\begin{equation*}
%\alpha_{q+1,q+1,k}
%=\mathsf{E}\left[\left.f(X^x_{q+1})\phi_{k}(\xi_{q+1})\right|\mathcal G_q\right].
%\end{equation*}
%By the Markov property of $(X^x_{l})$, it holds $\mathsf{E}[f(X^x_{q+1})|\mathcal{G}_q] = \mathsf{E}[f(X^x_{q+1})| X_q^x]$. Using Markov property together with conditioning on $\mathcal{G}_{q+1}$, we obtain
%\[
%\alpha_{q+1,q+1,k}=\mathsf{E}\left[\left.f(X^x_{q+1})\phi_{k}(\xi_{q+1})\right|X^x_{q}\right]
%=a_{q+1,q+1,k}(X^x_{q})
%\]
%where for $y \in \rset^d$
%\[
%a_{q+1,q+1,k}(y) = \mathsf{E}\left[f(X^y_{q,q+1})\phi_k\left(\xi_{q+1}\right)\right]
%\]
%We thus arrive at
%\begin{align}
%\label{berm:sig_X}
%f(X^x_{q+1})=\mathsf{E}\left[\left.f(X^x_{q+1})\right|X^x_{q}\right]+\sum_{k\geq1}a_{q+1,q+1,k}(X^x_{q})\phi_{k}(\xi_{q+1}),
%\end{align}
which is the required statement in the case $j=q$.
Now assume $j<q$.
The random variable
$\mathsf{E}\left[\left.f(X^x_{q+1})\right|X^x_{q}\right]$
is square integrable and has the form
$g(X^x_{q})$,
hence the induction hypothesis applies, and we get
\begin{equation}\label{eq:28082017a1}
\mathsf{E}\left[\left.f(X^x_{q+1})\right|X^x_{q}\right]=\mathsf{E}\left[\left.f(X^x_{q+1})\right|X^x_{j}\right]+\sum_{k\geq1}\sum_{l=j+1}^{q}a_{q+1,l,k}(X^x_{l-1})\phi_{k}(\xi_{l})
with
a_{q+1,l,k}(X^x_{l-1}) &= \mathsf{E}\left[\left.\mathsf{E}\left[\left.f(X^x_{q+1})\right|\mathcal{G}_{q}\right]\phi_{k}(\xi_{l})\right|\mathcal{G}_{l-1}\right]
 = \mathsf{E}\left[\left.f(X^x_{q+1})\phi_{k}(\xi_{l})\right|X^x_{l-1}\right].
where for $y \in \rset^d$,
a_{q+1,l,k}(y) = \mathsf{E}\left[f(X^y_{l-1,q+1})\phi_{k}(\xi_{l})\right]
Formulas \eqref{eq:1-step-decomp}
and~\eqref{eq:28082017a1} conclude the induction step for $p = q+1$ and hence the proof.
